An Interpretable Machine Learning Framework for Modeling High-Resolution Spectroscopic Data\footnote{Open source code at \url{https://github.com/gully/blase}}

### 2022-10-03 paper/paper1/ms.tex
Add some text from comments by Phill Cargile and Hajime Kawahara, thank you!!

For example, exoplanet cross-correlation spectroscopy \citep{2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.

@@

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'' or ``autograd'',][]{2015arXiv150205767G, 2016PhDT.......317M} and its affiliated backpropagation algorithm \citep{kelley1960,Linnainmaa1976TaylorEO, 1986Natur.323..533R, 1990JGCD...13..926D} has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{morphine} \citep{2021ApJ...907...40P,2021JOSAB..38.2465W} \edit1{and $\partial$\texttt{lux} \citep{10.1117/12.2629774}}, \edit1{stellar surface modeling with \texttt{starry} \citep{2021AJ....162..123L}}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \citep{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

@@

We place mild demands on the precomputed spectra. They should have sporadic regions of discernable continuum devoid of lines, and the continuum shape should vary smoothly in wavelength. The spectral lines or pseudo-lines should be resolved (and not sub-sampled). We suspect most stellar spectra meet these criteria, except for the coolest M dwarfs and brown dwarfs. The method can handle spectra without discernable continua, with some fine-adjustments to the preprocessing steps. We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$.  \edit1{The buffers are needed for two reasons: 1) to include the influence of spectra lines whose centers lie just outside the region-of-interest, but whose wings are broad enough to affect the neighboring pixels; and 2) to account for astrophysical radial velocity shifts that may send line centers into- and out-of- the region of interest.}  We set the value of the buffers to account for plausible radial velocity and rotational broadening of real stars. A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

\subsection{Goodness of fit metric and gradients}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes. Instead, we tune the parameters of the model, starting from these coarse initial values. This model has between $N_{\mathrm{lines}}\times 1$ and $N_{\mathrm{lines}}\times 4$ free parameters, depending on how many of the 4 physical line properties you wish to tune.  We default to fitting three parameters per line, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary. We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{S}_{\rm clone}$ but has no notion of per-pixel noise since the precomputed synthetic spectrum has no uncertainty:

@@

Most autodiff-aware \texttt{convolve} operators act in pixel-space, approximating kernels as numerically sampled functions.  There exists a special exponential spectral sample spacing that allows the convolution operators to work out of the box for rotational broadening.  The design of \emph{blas\'e} permits the stellar and telluric models to be re\"evaluated at any wavelength coordinate vector, provided that it adequately samples the underlying lines.  We, therefore, change the sampling from the native stellar and telluric wavelength coordinate grids, $\bm{\lambda}_S$ and $\bm{\lambda}_T$, to this special exponentially sampled wavelength grid, denoted with the subscript $Z$:

@@

Equation \ref{eqn_final_model} has $\sim21,000$ tunable parameters from the star, $\sim9,000$ tunable parameters from the Earth's atmosphere, plus $v\sin{i}$ and $RV$.  That adds up to about $30,002$ model parameters.  The resolving power may also be treated as tunable if it is not known or varies slightly with \emph{e.g.} seeing, slit-or-fiber illumination, or instrumental configuration: $30,003$.

@@

>>>>

For example, exoplanet cross-correlation spectroscopy \citep{2010Natur.465.1049S,2012Natur.486..502B,2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.

@@

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'' or ``autograd'',][]{2015arXiv150205767G, 2016PhDT.......317M} and its affiliated backpropagation algorithm \citep{kelley1960,Linnainmaa1976TaylorEO, 1986Natur.323..533R, 1990JGCD...13..926D} has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{morphine} \citep{2021ApJ...907...40P,2021JOSAB..38.2465W} \edit1{and $\partial$\texttt{lux} \citep{10.1117/12.2629774}}, \edit1{stellar surface modeling with \texttt{starry} \citep{2021AJ....162..123L}}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \citep{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$100 tunable parameters that describe the fundamental physical properties controlling a \edit2{stratified} brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

@@

We place mild demands on the precomputed spectra. They should have sporadic regions of discernable continuum \edit2{(or pseudocontinuum)} devoid of lines, and the continuum shape should vary smoothly in wavelength. The spectral lines or pseudo-lines should be resolved (and not sub-sampled). Many precomputed synthetic stellar spectra will meet these criteria. \edit2{For those spectra that do not meet these criteria, the method should still work with some additional fine-tuning of the pre-processing steps that follow.  M dwarfs in the red-optical, solar-like stars in the blue visible, and brown dwarfs at virtually any wavelength may fall into this category.}

We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$.  \edit1{The buffers are needed for two reasons: 1) to include the influence of spectral lines whose centers lie just outside the region-of-interest, but whose wings are broad enough to affect the neighboring pixels; and 2) to account for astrophysical radial velocity shifts that may send line centers into- and out-of- the region of interest.}  We set the value of the buffers to account for plausible radial velocity and rotational broadening of real stars. A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

\subsection{Goodness of fit metric and gradients}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes. Instead, we tune the parameters of the model, starting from these coarse initial values. This model has between $N_{\mathrm{lines}}\times 1$ and $N_{\mathrm{lines}}\times 4$ free parameters, depending on how many of the 4 physical line properties you wish to tune.  \edit2{The treshold-based line-finding algorithm performs very well at finding the line center positions, so in practice we can fix the line center wavelengths.}  We therefore default to fitting three parameters per line, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary. We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{S}_{\rm clone}$ but has no notion of per-pixel noise since the precomputed synthetic spectrum has no uncertainty:

@@

Most autodiff-aware \texttt{convolve} operators act in pixel-space, approximating kernels as numerically sampled functions.  There exists a special exponential spectral sample spacing that allows the convolution operators to work out of the box for rotational broadening.  The design of \emph{blas\'e} permits the stellar and telluric models to be reevaluated at any wavelength coordinate vector, provided that it adequately samples the underlying lines.  We, therefore, change the sampling from the native stellar and telluric wavelength coordinate grids, $\bm{\lambda}_S$ and $\bm{\lambda}_T$, to this special exponentially sampled wavelength grid, denoted with the subscript $Z$:

@@

Equation \ref{eqn_final_model} has $\sim21,000$ tunable parameters from the star, $\sim9,000$ tunable parameters from the Earth's atmosphere, plus $v\sin{i}$ and $RV$.  That adds up to about $30,002$ model parameters.  The resolving power may also be treated as tunable if it is not known or varies slightly with \emph{e.g.} seeing, slit-or-fiber illumination, or instrumental configuration: $30,003$.  \edit2{The pseudocontinuum polynomial term $\mathsf{P}$ can also be tuned to refine the continuum placement initialized at the pre-processing stage, yielding an additional 5-15 parameters, depending on the choice of polynomial order.}

@@

\subsection{Sky emission lines and wavelength calibration}
\edit2{Night sky emission lines---arising from OH, for example---add emission line spikes on top of the astronomical target spectrum of interest.  These lines get subtracted in most pipeline packages, but in some cases a more careful treatment may be desirable.  In HPF, a sky reference fiber points towards blank sky, needing some careful calibration before subtraction from the target fiber \citep{2022JOSS....7.4302G}.  Remote sensing applications may wish to measure such emission as the primary science of interest.  It would be straightforward for \emph{blas\'e} to handle emission lines, following the augmentation for emission lines in Section \ref{gpuConsiderations}.  The main benefit over conventional sky subtraction methods may be the ability to disentangle the underlying spectrum at super-resolution.  A predictive model for sky emission could be built-up over time, building data-driven corrections to theoretical models.}

\edit2{The well-known center wavelengths of these lines could be used for additional wavelength calibration, as is already done in IGRINS \citep{jaejoonlee16}.  An imperfect instrumental wavelength calibration could be diagnosed based on departure of the predicted versus realized wavelength center positions, enabling a relatively fast and intervention-free mitigation strategy.  Hypothetically, a prescription for the wavelength calibration could be built directly into \emph{blas\'e}, with the coefficents for a Polynomial re-routing the realized wavelength center locations from their known wavelength centers.
}
